---
title: "Chills terms - visualisations"
format:
  html:
    code-fold: true
    code-summary: "Show the code"
---

# Aim

- Take the data directly from the google sheets (downloaded as excel sheet for offline processing, `Chills_Definitions.xlsx`).

- Process the definitions into network visualisation to show a pattern between the terminology used, the affective characteristics, and the physiological characteristics.

## Preliminaries

### Load libraries

```{r}
#| id: libraries
#| warning: false
#| message: false
library(tidyr)
#library(tidytext)
library(dplyr)
library(ggplot2)
library(stringr)

```

### Read data

```{r}
#| id: read-data
#| warning: false
#| message: false
#| output: asis
d <- readxl::read_excel("data/Chills_Definitions.xlsx")
d$Year<-stringr::str_remove(d$Year, "\\.0$") # remove trailing .0 from years
d$Page<-stringr::str_remove(d$Page, "\\.0$") # remove trailing .0 from pages
print(knitr::kable(head(d,3)))
```


### Recode definitions

1. Collapse some rare **terminology** categories (such as "Chill(ing sound)" or "Tingling" or "Shivers") into "Other"
2. Collapse terms for **affects**.
3. Collapse terms for **physiology**.
4. Show amount of misssing data in each category.

```{r}
#| id: recode
#| warning: false
#| output: asis

## 1. Terminologies simplified
d$Terminology <- recode(d$Terminology,
  "Chill(ing sound)" = "Other",
  "Horripilation" = "Other",
  "Tingling" = "Other",
  "Shivers" = "Other",
  "Thrill / Chill" = "Thrill",
  "Chill / Frisson Chill" = "Chill",
  "Chill / Piloerection" = "Chill",
  "Chill / Frisson" = "Frisson",
  "Chill / Thrill" = "Thrill"
)
# also explictly code those rows that have NA in Terminology as "Not Defined" so they are included in the visualisations
d$Terminology[is.na(d$Terminology)]<- "Not Defined"
print(knitr::kable(as.data.frame(table(d$Terminology)), col.names = c("Terminology", "N")))

## 2. Affect
d$Affect <- recode(d$Affect,
  "Affective change" = "Affect variant",
  "Aversive" = "Affect variant",
  "Embodied" = "Other",
  "Emotional" = "Affect variant",
  "Excitement" = "Affect variant",
  "Euphoric" = "Affect variant",
  "Meaningful" = "Other",
  "Merging with music" = "Other",
  "Negative" = "Affect variant",
  "Sexual arousal" = "Affect variant",
  "Strong" = "Affect variant",
  "Subjective" = "Other",
  "Fear" = "Affect variant",
  "Motor-sensory ecstasy" = "Affect variant",
  "Unpleasant" = "Affect variant")

d$Affect[is.na(d$Affect)]<- "Not Defined"
print(knitr::kable(as.data.frame(table(d$Affect)), col.names = c("Affect", "N")))

## 3. Simplify Physiology
d$Physiology[is.na(d$Physiology)]<- "Not Defined"
print(knitr::kable(as.data.frame(table(d$Physiology)), col.names = c("Physiology", "N")))

## 4. Simplify Body location?
# rename the column, at least
names(d)[names(d) == "Body Location"] <- "Body_Location"
d$Body_Location[is.na(d$Body_Location)]<- "Not Defined"
print(knitr::kable(as.data.frame(table(d$Body_Location)), col.names = c("Body location", "N")))

## Summarise not defined in each category
cat_summary <- data.frame(
  Category = c("Terminology", "Affect", "Physiology", "Body Location"),
  Not_Defined_pct = c(sum(d$Terminology == "Not Defined")/nrow(d),
                        sum(d$Affect == "Not Defined")/nrow(d),
                        sum(d$Physiology == "Not Defined")/nrow(d),
                        sum(d$Body_Location == "Not Defined")/nrow(d))
)
print(knitr::kable(cat_summary,digits=2))

```

## Visualise as network graph

Visualise the co-occurrence of Terminology, Affect, and Physiology using a network graph.

Principles:
- The order of the nodes is now determined by frequency within each category. 
- There is a filter to remove rare terms (<3).
- "Not defined" is included as a node.
- Node positions can be made to be more organic, now rigid.

```{r}
#| id: final-network
#| warning: false
#| message: false
#| fig.width: 8
#| fig.height: 6

library(igraph)
library(ggraph)

# Select the terms of interest:
terms <- dplyr::select(d, Terminology, Affect, Physiology)

terminology_terms <- terms %>% pull(Terminology) %>% unique()

node_freq <- bind_rows(
  terms %>% select(term = Terminology),
  terms %>% select(term = Affect),
  terms %>% select(term = Physiology)
) %>%
  filter(!is.na(term)) %>%
  count(term)

freq_threshold <- 3  # find suitable threshold

# Make sure to keep terms that are related to the nodes, so make them unique here. 
node_freq <- bind_rows(
  terms %>% transmute(term = paste0(Terminology, " [T]")),
  terms %>% transmute(term = paste0(Affect,      " [A]")),
  terms %>% transmute(term = paste0(Physiology,  " [P]"))
) %>%
  filter(!is.na(term)) %>%
  count(term) %>%
  filter(n >= freq_threshold)   # drop rare terms

edges <- bind_rows(
  terms %>% mutate(from = paste0(Terminology, " [T]"),
                   to   = paste0(Affect,      " [A]")) %>%
            select(from, to),
  terms %>% mutate(from = paste0(Affect,      " [A]"),
                   to   = paste0(Physiology,  " [P]")) %>%
            select(from, to)
) %>%
  filter(!is.na(from), !is.na(to)) %>%
  filter(from %in% node_freq$term, to %in% node_freq$term) %>%  # drop edges to removed nodes
  count(from, to, name = "weight")


g <- graph_from_data_frame(edges) %>%
  tidygraph::as_tbl_graph() %>%
  left_join(node_freq, by = c("name" = "term")) %>%
  mutate(
    column = case_when(
      stringr::str_ends(name, "\\[T\\]") ~ "Terminology",
      stringr::str_ends(name, "\\[A\\]") ~ "Affect",
      stringr::str_ends(name, "\\[P\\]") ~ "Physiology"
    ),
    # Strip the suffix for display
    label = stringr::str_remove(name, " \\[.\\]$")
  )

# Centred y coordinates: rank within group, then subtract the group mean
g <- g %>%
  mutate(
    x = case_when(
      column == "Terminology" ~ 0,
      column == "Affect"      ~ 1,
      column == "Physiology"  ~ 2
    ),
    y = {
      ranks <- ave(seq_along(name), column, FUN = seq_along)
      group_min  <- ave(ranks, column, FUN = min)
      group_max  <- ave(ranks, column, FUN = max)
      # Rescale to [-1, 1]; guard against single-node groups (max == min)
      ifelse(
        group_max == group_min,
        0,
        2 * (ranks - group_min) / (group_max - group_min) - 1
      )
    }
  )


g <- g %>%
  mutate(
    x = case_when(
      column == "Terminology" ~ 0,
      column == "Affect"      ~ 1,
      column == "Physiology"  ~ 2
    )
  ) %>%
  arrange(column, n) %>%   # sort by frequency within group *was desc(n)
  mutate(
    y = {
      ranks     <- ave(seq_along(name), column, FUN = seq_along)
      group_min <- ave(ranks, column, FUN = min)
      group_max <- ave(ranks, column, FUN = max)
      ifelse(
        group_max == group_min,
        0,
        2 * (ranks - group_min) / (group_max - group_min) - 1
      )
    }
  )

## Add labels
col_labels <- data.frame(
  x     = c(0, 1, 2),
  y     = rep(1.15, 3),          # just above the top node at y = 1
  label = c("Terminology", "Affect", "Physiology")
)

## Actual figure
fig1 <- g %>%
  ggraph(layout = "manual", x = x, y = y) +
  geom_edge_link(aes(width = weight,color=weight), alpha = 0.4) +
  geom_node_point(aes(size = n, colour = column), alpha = 1.0) +
  geom_node_text(aes(label = label), repel = TRUE, size = 5) +
  scale_edge_colour_gradientn(colours = c("grey80", "grey50", "grey20"), values  = c(0, 0.3, 1)) +
  scale_x_continuous(expand = expansion(add = 0.5)) +  # add padding left & right
  scale_edge_width(range = c(0.5, 3)) +
  geom_text(
    data    = col_labels,
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    fontface    = "bold",
    size        = 5
  ) +
  scale_size_continuous(range = c(3, 12), name = "Frequency") +
  scale_colour_manual(
    values = c("Terms" = "steelblue",
               "Affect"      = "tomato",
               "Physiology"  = "seagreen"),
    na.value = "steelblue"
  ) +
  guides(
    size   = guide_legend(),   # keep node size
    colour = "none",           # drop colour
    edge_width = "none",        # drop edge width
    edge_colour  = "none"        # drop edge colour legend
  ) +
  theme_void()+
  theme(legend.position = "bottom")

print(fig1)

```
